From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-04 Pi+/paper/reversible1.tex
scale math
minor fixes in sec 2

\medskip\noindent The language of types is built from the empty type ($\zerot$), the unit type
($\onet$) containing just one value~$\Acon{tt}$, the sum type ($+$) containing values of the form $\inlv{v}$ and
$\inrv{v}$, and the product ($\times$) type containing pairs of values $(v_1,v_2)$.
%
% A natural candidate for a semantic foundation for reversible programming languages is the notion of type
% isomorphism. Indeed, the type isomorphisms among finite types are sound and complete for all permutations on finite
% types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
% circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980} and the extension with recursive
% types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
% language~\cite{James:2012:IE:2103656.2103667,rc2011}.

To see how this language is expresses reversible circuits, we present a few examples. First it is possible to directly
mimic the \verb|qasm|-perspective by defining types that describe sequences of booleans. We use the type
$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true} and
$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define using
the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus the
type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

@@

All of this is only half the story as the correspondence between the syntactic commutative rig (the syntax of $\PiLang$)
and permutations on finite sets includes \emph{coherence conditions} that identify different syntactic representations
of the same permutation~\cite{laplaza72,Carette2016}. These coherence conditions will be reflected in a second level of $\PiLang$
syntax as level-2 combinators whose types are of the form $c_1 \isotwo c_2$ for appropriate $c_1$ and $c_2$ of the same
level-1 type $A \isoone B$. For example, we have:

@@

\begin{align*}
  \denot{\zerot}     & = \emptyset                  \\
  \denot{\onet}      & = \{ \Acon{tt} \}            \\
  \denot{A + B}      & = \denot{A} \uplus \denot{B} \\
  \denot{A \times B} & = \denot{A} \times \denot{B}
\end{align*}

@@

\begin{multicols}{2}
%%  \begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
\[\begin{array}[t]{rcl}
  \denot{\identlp}     ~      (\inlv{v})           & = & v                                   \\
  \denot{\identrp}      ~      v                    & = & \inlv{v}                            \\
  \denot{\swapp}        ~      (\inlv{v})           & = & \inrv{v}                            \\
  \denot{\swapp}        ~      (\inrv{v})           & = & \inlv{v}                            \\
  \denot{\assoclp}      ~      (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
  \denot{\assoclp}      ~      (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
  \denot{\assoclp}      ~      (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
  \denot{\assocrp}      ~      (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
  \denot{\assocrp}      ~      (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
  \denot{\assocrp}      ~      (\inrv{v})           & = & \inrv{(\inrv{v})}
\end{array}\]

\[\begin{array}[t]{rcl}
  \denot{\identlt}      ~      (\Acon{tt} , v)      & = & v                                   \\
  \denot{\identrt}      ~      v                    & = & (\Acon{tt} , v)                     \\
  \denot{\swapt}        ~      (v_1 , v_2)          & = & (v_2 , v_1)                         \\
  \denot{\assoclt}      ~      (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
  \denot{\assocrt}      ~      ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
  \denot{\dist}         ~      (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
  \denot{\dist}         ~      (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
  \denot{\factor}       ~      (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
  \denot{\factor}       ~      (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)
\end{array}\]

@@

  \[\begin{array}{rcl}
  \denot{\idc}          ~      v                    & = & v                                   \\
  \denot{(c_1 \fatsemi c_2)} ~ v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
  \denot{(c_1 \oplus c_2)}   ~ (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
  \denot{(c_1 \oplus c_2)}   ~ (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
  \denot{(c_1 \otimes c_2)}  ~ (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
  \end{array}\]

@@

\begin{theorem}\label{thm:semone}
  The denotational semantics is sound in the following sense:
  \begin{itemize}
    \item For every level-1 combinator $c : A \isoone B$, we have that $\denot{c}$ is a bijection between $\denot{A}$ and $\denot{B}$.
      \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
        combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using the
        extensional equivalence of functions.
  \end{itemize}
\end{theorem}
%% proof?

>>>>

\medskip\noindent The language of types is built from the empty type ($\zerot$), the unit type
($\onet$) containing just one value~$\Acon{tt}$, the sum type ($+$) containing values of the form $\inlv{v}$ and
$\inrv{v}$, and the product type ($\times$) containing pairs of values $(v_1,v_2)$.
%
% A natural candidate for a semantic foundation for reversible programming languages is the notion of type
% isomorphism. Indeed, the type isomorphisms among finite types are sound and complete for all permutations on finite
% types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
% circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980} and the extension with recursive
% types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
% language~\cite{James:2012:IE:2103656.2103667,rc2011}.

To see how this language expresses reversible circuits, we present a few examples. First it is possible to directly
mimic the \verb|qasm|-perspective by defining types that describe sequences of booleans. We use the type
$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true} and
$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define using
the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus the
type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

@@

All of this is only half the story however as the correspondence between the syntactic commutative rig (the syntax of $\PiLang$)
and permutations on finite sets includes \emph{coherence conditions} that identify different syntactic representations
of the same permutation~\cite{laplaza72,Carette2016}. These coherence conditions are collected in a second level of $\PiLang$
syntax as level-2 combinators whose types are of the form $c_1 \isotwo c_2$ for appropriate $c_1$ and $c_2$ of the same
level-1 type $A \isoone B$. For example, we have:

@@

\begin{center}
{\scalebox{\scalef}{$%
\begin{array}{rcl}
  \denot{\zerot}     & = & \emptyset                  \\
  \denot{\onet}      & = & \{ \Acon{tt} \}            \\
  \denot{A + B}      & = & \denot{A} \uplus \denot{B} \\
  \denot{A \times B} & = & \denot{A} \times \denot{B}
\end{array}$}}
\end{center}

@@

\begin{multicols}{2}
%%  \begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
{\scalebox{\scalef}{$%
\begin{array}[t]{rlcl}
  \denot{\identlp}     &      (\inlv{v})           & = & v                                   \\
  \denot{\identrp}     &      v                    & = & \inlv{v}                            \\
  \denot{\swapp}       &     (\inlv{v})           & = & \inrv{v}                            \\
  \denot{\swapp}       &      (\inrv{v})           & = & \inlv{v}                            \\
  \denot{\assoclp}     &      (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
  \denot{\assoclp}     &      (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
  \denot{\assoclp}     &      (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
  \denot{\assocrp}     &      (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
  \denot{\assocrp}     &      (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
  \denot{\assocrp}     &      (\inrv{v})           & = & \inrv{(\inrv{v})}
\end{array}$}}

{\scalebox{\scalef}{$%
\begin{array}[t]{rlcl}
  \denot{\identlt}      &      (\Acon{tt} , v)      & = & v                                   \\
  \denot{\identrt}      &      v                    & = & (\Acon{tt} , v)                     \\
  \denot{\swapt}        &      (v_1 , v_2)          & = & (v_2 , v_1)                         \\
  \denot{\assoclt}     &      (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
  \denot{\assocrt}     &      ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
  \denot{\dist}       &      (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
  \denot{\dist}        &      (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
  \denot{\factor}       &      (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
  \denot{\factor}      &      (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)
\end{array}$}}

@@

\begin{center}
{\scalebox{\scalef}{$%
\begin{array}{rlcl}
  \denot{\idc}         &      v                    & = & v                                   \\
  \denot{(c_1 \fatsemi c_2)} & v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
  \denot{(c_1 \oplus c_2)}   & (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
  \denot{(c_1 \oplus c_2)}   & (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
  \denot{(c_1 \otimes c_2)}  & (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
  \end{array}
$}}
\end{center}

@@

\begin{theorem}\label{thm:semone}
  The denotational semantics is sound in the following sense:
  \begin{itemize}
    \item For every level-1 combinator $c : A \isoone B$, we have that $\denot{c}$ is a bijection between $\denot{A}$ and $\denot{B}$.
      \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
        combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using
        extensional equivalence of functions.
  \end{itemize}
\end{theorem}
%% proof?
